Runtime Feedback in a Meta-Tracing JIT for Efficient Dynamic Languages

### 2011-04-01 talk/icooolps2011/paper.tex
small tweaks, add a citation, reword a paragraph

\authorinfo{Carl Friedrich Bolz$^a$ \and Antonio Cuni$^a$ \and Maciej Fijałkowski$^b$ \and Michael Leuschel$^a$ \and \\
            Samuele Pedroni$^c$ \and Armin Rigo$^a$}
           {$^a$Heinrich-Heine-Universität Düsseldorf, STUPS Group, Germany

            $^b$merlinux GmbH, Hildesheim, Germany

            $^c$Open End, Göteborg, Sweden
           }
           {cfbolz@gmx.de \and anto.cuni@gmail.com \and fijal@merlinux.eu \and
           leuschel@cs.uni-duesseldorf.de \and samuele.pedroni@gmail.com \and arigo@tunes.org}

@@

Meta-tracing JITs can be applied to a variety of different
languages without explicitly encoding language semantics into the compiler. So
far, they lacked a way to feed back runtime information into the
compiler, which restricted their performance. In this paper we describe the
flexible mechanisms in PyPy's meta-tracing JIT that can be used to control runtime feedback in language-specific ways. These mechanisms are flexible
enough to implement classical VM techniques such as maps and polymorphic inline
caches.

@@

One of the hardest parts of implementing a dynamic language efficiently is to
optimize its object model. This is made harder by the fact that many recent
languages such as Python, JavaScript or Ruby have rather complex core object
semantics. For them, implementing just an interpreter is already a complex
task. Implementing them efficiently with a just-in-time compiler (JIT) is
extremely challenging, because of their many corner-cases.

It has long been an objective of the partial evaluation community to
automatically produce compilers from interpreters. There has been a recent
renaissance of this idea around the approach of tracing just-in-time
compilers. A number of projects have attempted this approach. SPUR \cite{bebenita_spur:_2010} is
a tracing JIT for .NET together with a JavaScript implementation in C\#. PyPy
\cite{armin_rigo_pypys_2006} contains a tracing JIT for Python (a restricted
subset of Python). This JIT is then used to trace a number of languages
implementations written in RPython. A number of other experiments in this
directions were done, such as an interpreter for Lua in JavaScript, which is run
on and optimized with a tracing JIT for JavaScript
\cite{yermolovich_optimization_2009}.

These projects have in common that they work one meta-level down, providing a tracing JIT for the implementation
language used to implement the dynamic language, and not for the dynamic language itself.
The tracing JIT then will trace through the object model of the dynamic
language implementation. This makes the object model transparent to the tracer
and its optimizations. Therefore the semantics of the dynamic language does not
have to be replicated in a JIT. We call this approach \emph{meta-tracing}.
Another commonality of these approaches is that they allow some annotations (or
hints) in the dynamic language implementation to guide the meta-tracer. This
makes the process not completely automatic but can give good speedups over
bare meta-tracing.

@@

Conceptually the significant speed-ups that can be achieved with
dynamic compilation depend on feeding into compilation and exploiting
values observed at runtime that are slow-varying in practice. To exploit the
runtime feedback, the implementation code and data structures need to be
structured so that many such values are at hand. The hints that we present allow
exactly to implement such feedback and exploitation in a meta-tracing context.

@@

@@

To be able to do this recording, VMs with a
tracing JIT typically contain an interpreter. After a user program is
started the interpreter is used; only the most frequently executed paths through the user
program are turned into machine code. The tracing JIT tries to produce traces
that correspond to loops in the traced program, but most tracing JITs now also
have support for tracing non-loops \cite{andreas_gal_incremental_2006}.

@@

Tracing through the execution of an interpreter has many advantages. It makes
the tracer, its optimizers and backends reusable for a variety of languages. The
language semantics do not need to be encoded into the JIT. Instead the tracer
just picks them up from the interpreter.

@@

For space reasons we cannot perform a full evaluation here, but still want to
present some benchmark numbers. We chose to present two benchmarks, a port of
the classical Richards benchmark in Python and a Python version of the Telco
decimal benchmark\footnote{\texttt{http://speleotrove.com/decimal/telco.html}},
using a pure Python decimal floating point implementation. The results we see in
these two benchmarks seem to repeat themselves in other benchmarks using
object-oriented code, for purely numerical algorithms the speedups are a lot
lower.

The benchmarks were run on an otherwise idle Intel Core2 Duo P8400 processor
with 2.26 GHz and 3072 KB of cache on a machine with 3GB RAM running Linux
2.6.35. We compared the performance of various Python implementations on the
benchmarks. As a baseline, we used the standard Python implementation in C,
CPython 2.6.6\footnote{\texttt{http://python.org}}, which uses a bytecode-based
interpreter. We compare it against four versions of PyPy's Python interpreter,
all of them with JIT enabled. The PyPy baseline does not enable maps or type
versions. Then we have a version each where maps and versions are enabled alone
and finally a version with both.

All benchmarks were run 50 times in the same process, to give the JIT time to
produce machine code. The arithmetic mean of the times of the last 30 runs were
used as the result. The errors were computed using a confidence interval with a
95\% confidence level \cite{georges_statistically_2007}. The results are
reported in Figure~\ref{fig:times}.

Versioned types speed up both benchmarks by a significant factor of around 7.
The speed advantage of maps alones is a lot less clear. Maps also have a memory
advantage which we did not measure here. By themselves, maps improved the
Richards benchmark slightly, but made the Telco benchmark slower. Enabling both
maps and versioned types together yields a significant improvement over just
versioned types for Richards. XXX good explanation. For Telco, enabling both
does not change much.

\pedronis{XXX radical idea, given that there may be no space to discuss the subtle points and that this is really about showing that we can enable such mechanisms and there is already literature that shows that/how they work, to just consider the benchmarks with no maps and no versions and with both?}

\begin{figure}
{\footnotesize
\begin{center}
\begin{tabular}{|l|r|r|}
\hline
        &richards[ms]       &telco[ms] \\
\hline
CPython &357.79 $\pm$ 1.32     &1209.67 $\pm$ 2.20\\
speedup &1.00 $\times$ &1.00 $\times$\\
\hline
JIT baseline    &421.87 $\pm$ 0.48     &738.18 $\pm$ 3.29\\
speedup &0.85 $\times$ &1.64 $\times$\\
\hline
JIT map &382.88 $\pm$ 4.40     &834.19 $\pm$ 4.91\\
speedup &0.93 $\times$ &1.45 $\times$\\
\hline
JIT version     &49.87 $\pm$ 0.29      &157.88 $\pm$ 1.79    \\
speedup &7.17 $\times$ &7.66 $\times$\\
\hline
JIT full        &17.89 $\pm$ 1.15      &153.48 $\pm$ 1.86    \\
speedup &20.00 $\times$        &7.88 $\times$\\
\hline
\end{tabular}
\end{center}
}
\caption{Benchmark Results}
\label{fig:times}
\end{figure}

@@

Partial evaluation \cite{jones_partial_1993} tries to automatically transform
interpreters into compilers using the second futamura projection
\cite{futamura_partial_1999}. Given that classical partial evaluation works
strictly ahead of time, it inherently cannot support runtime feedback.

@@

Somewhat relatedly, the proposed ``invokedynamic'' bytecode
\cite{rose_bytecodes_2009} that will be added to the JVM is supposed to make the
implementation of dynamic languages on top of JVMs easier. The bytecode gives
the language implementor control over how the JIT optimizes the language's
features and when optimized code needs to be deoptimized. XXX

We already explored promotion in other context, such as earlier versions of
PyPy's JIT \cite{armin_rigo_jit_2007} as well as a Prolog partial evaluator
\cite{carl_friedrich_bolz_towards_????}. Promotion is quite similar to
(polymorphic) inline caching and runtime type feedback techniques which were
first used in Smalltalk \cite{deutsch_efficient_1984} and SELF
\cite{hoelzle_optimizing_1991,hoelzle_optimizing_1994} implementations.
Promotion is more general because any information can be cached in line, not
just classes of method receivers.

@@

%jruby used versions at some point, jvm-l mailing list discusses them

\section{Conclusion}

In this paper we presented two hints that can be used in the source code of an
interpreter written with PyPy. They give control over runtime feedback and
optimization to the language implementor. They are expressive enough for
building well-known virtual machine optimization techniques, such as maps and
inline caches. We believe that they are flexible enough to express a wide
variety of language semantics efficiently.

>>>>

\authorinfo{Carl Friedrich Bolz \and XXX}
           {Heinrich-Heine-Universität Düsseldorf, STUPS Group, Germany
           }
           {cfbolz@gmx.de \and XXX}

@@

@@

One of the hardest parts of implementing a dynamic language efficiently is to
optimize its object model. This is made harder by the fact that many recent
languages such as Python, JavaScript or Ruby have rather complex core object
semantics. For them, even implementing just an interpreter is already a complex
task. Implementing them efficiently with a just-in-time compiler (JIT) is
extremely challenging, because of their many corner-cases.

It has long been an objective of the partial evaluation community to
automatically produce compilers from interpreters. There has been a recent
renaissance of this idea around the approach of tracing just-in-time
compilers. A number of projects have attempted this approach. SPUR \cite{bebenita_spur:_2010} is
a tracing JIT for .NET together with a JavaScript implementation in C\#. PyPy
\cite{armin_rigo_pypys_2006} contains a tracing JIT for RPython \cite{davide_ancona_rpython:_2007} (a restricted
subset of Python). This JIT is then used to trace a number of languages
implementations written in RPython. A number of other experiments in this
directions were done, such as an interpreter for Lua in JavaScript, which is run
on and optimized with a tracing JIT for JavaScript
\cite{yermolovich_optimization_2009}.

These projects have in common that they work one meta-level down, providing a tracing JIT for the
language used to implement the dynamic language, and not for the dynamic language itself.
The tracing JIT then will trace through the object model of the dynamic
language implementation. This makes the object model transparent to the tracer
and its optimizations. Therefore the semantics of the dynamic language does not
have to be replicated in a JIT. We call this approach \emph{meta-tracing}.
Another commonality of these approaches is that they allow some annotations (or
hints) in the dynamic language implementation to guide the meta-tracer. This
makes the process not completely automatic but can give good speedups over
bare meta-tracing.

@@

% XXX: paragraph rephrased by anto; feel free to pick the one you like best

Conceptually, it is possible to achieve significant speed-ups by feeding into
the compiler some information that is observed at runtime: in particular, if
there are values which vary very slowly, it is possible to compile multiple
specialized versions of the same code, one for each actual value.  To exploit
the runtime feedback, the implementation code and data structures need to be
structured so that many such slow-varying values are at hand. The hints that
we present allow exactly to implement such feedback and exploitation in a
meta-tracing context.

% Conceptually the significant speed-ups that can be achieved with
% dynamic compilation depend on feeding into compilation and exploiting
% values observed at runtime that are slow-varying in practice. To exploit the
% runtime feedback, the implementation code and data structures need to be
% structured so that many such values are at hand. The hints that we present allow
% exactly to implement such feedback and exploitation in a meta-tracing context.

@@

The advantage of this approach is that writing an interpreter is much easier
and less error prone than manually writing a JIT compiler.  Similarly, writing
in a high level language such as RPython is easier than writing in C.

@@

To be able to do this recording, VMs with a
tracing JIT typically contain an interpreter. After a user program is
started the interpreter is used; only the most frequently executed paths through the user
program are turned into machine code. The tracing JIT tries to produce traces
that correspond to loops in the traced program, but most tracing JITs now also
have support for tracing non-loops \cite{XXX}.

@@

Tracing through the execution of an interpreter has many advantages. It makes
the tracer, its optimizers and backends reusable for a variety of languages. The
language semantics do not need to be encoded into the JIT. Instead the tracer
just picks them up from the interpreter. XXX mention disadvantage of long
traces?

@@

@@

Partial evaluation \cite{jones_partial_1993} tries to automatically transform
interpreters into compilers using the second futamura projection
\cite{futamura_partial_1999}. XXX

@@

Somewhat relatedly, the proposed ``invokedynamic'' bytecode
\cite{rose_bytecodes_2009} that will be added to the JVM is supposed to make the
implementation of dynamic languages on top of JVMs easier. The bytecode gives
the language implementor control over how the JIT optimizes the language's
features. XXX

%We already explored promotion in other context, such as earlier examples of
%promotion \cite{carl_friedrich_bolz_towards_????}, \cite{armin_rigo_jit_2007}

\cite{chambers_efficient_1989} maps
\cite{hoelzle_optimizing_1994} Type feedback
\cite{hoelzle_optimizing_1991} PICs
find the web page of V8 about maps

@@

jruby used versions at some point, jvm-l mailing list discusses them

\section{Conclusion and Next Steps}

In this paper we presented two hints that can be used in the source code of an
interpreter written with PyPy. They are used to influence what the optimizer
does with the trace. We also showed how a small but still relevant dynamic
object model can use these hints to no longer use any dictionary lookups after
tracing. Instead a number of guards are inserted into the trace to check whether
the assumptions about the objects are still true. This makes operations on
objects seriously faster.

XXX
